Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-05-29 doc/network.tex
sentence
simplify text

Here, we investigate two methods for correcting for temporal bias in a contact tracing data (see Methods for details).
We refer to the first method as the population-level method as it relies on the observed distribution aggregated across the entire population.
When an epidemic is growing exponentially, the right-censoring causes the observed generation-interval inverse exponentially weighted \eref{exp}; hence, we can ``undo'' the censoring by exponentially weighting the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
\begin{equation}
\tsup{g}{exp}(\tau) \propto c_t(\tau) \exp(r\tau),
\end{equation}
where $r$ is the exponential growth rate.
We refer to the second method as the individual-level method because it relies on individual contact information.
We model each infection as a non-homogeneous Poisson process arising from the infector; 
incorporating information about time of infection of an infector, time of infection of an infectee, and time since the beginning of an epidemic allows us to explicitly model the censoring process in the observed generation intervals.

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap} and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
\cite{trapman2016inferring} defined the empirical reproductive number by dividing the total number of infected individuals in generations $2$ to $k+1$ by the total number of infected individuals in generations $1$ to $k$, where the reference generation $k$ is defined to be the first generation that contains at least 75 infected individuals.
however, individuals in later generations are potentially subject to stronger local depletion and even global depletion effects.
Therefore, we expect our empirical estimate to better reflect the infection process during the exponential growth phase.
\swp{TODO: Compare Trapman vs our estimate of empirical R0 in Appendix; Trapman estimate is much lower.}

As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution (\eref{conditional}) are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects. \dc{Figure? I guess this is work in progress...}
Direct estimates based on contact tracing data severely underestimates the empirical estimates.
While both population- and individual-level corrections provide similar estimates to the empirical reproductive number,
population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
For smaller values of \RR, we expect the differences to become smaller. 
\swp{TODO: In Appendix, we present the same figure using Erlang-distributed latent periods, which better corresponds to Ebola. Overall, our qualitative conclusions do not change.}

>>>>

Here, we investigate two methods for correcting for temporal bias in a contact tracing data (see Methods for details).
We refer to the first method as the population-level method as it relies on the observed distribution aggregated across the entire population.
When an epidemic is growing exponentially, the right-censoring causes the observed generation-interval inverse exponentially weighted \eref{exp}; hence, we can ``undo'' the censoring by exponentially weighting the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
\begin{equation}
\tsup{g}{exp}(\tau) \propto c_t(\tau) \exp(r\tau),
\end{equation}
where $r$ is the exponential growth rate.
We refer to the second method as the individual-level method because it relies on individual contact information.
We model each infection as a non-homogeneous Poisson process arising from the infector; 
incorporating information about time of infection of an infector, time of infection of an infectee, and time since the beginning of an epidemic allows us to explicitly model the censoring process in the observed generation intervals.
For both methods, mean and coefficient of variation (CV) of generation-interval distributions are estimated by maximum likelihood; the inferred generation-interval distribution is then used to estimate the reproductive number $\RR$ from the observed growth rate $r$.

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap} and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution (\eref{conditional}) are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects. \dc{Figure? I guess this is work in progress...}
Direct estimates based on the observed generation intervals from contact tracing severely underestimates the empirical estimates.
While both population- and individual-level corrections provide similar estimates to the empirical reproductive number on average,
population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
For smaller values of \RR, we expect the differences to become smaller. 
\swp{TODO: In Appendix, we present the same figure using Erlang-distributed latent periods, which better corresponds to Ebola. Overall, our qualitative conclusions do not change.}
